Using occupancy theory and Bayesian models to estimate otter populations in southeastern Minnesota

### 2010-03-01 Presentation/Presentation.tex
finished presentation draft

\documentclass{beamer}
\usetheme{}
\title{Using occupancy theory and Bayesian models to estimate otter populations in southeastern Minnesota}
\author[Aing, Halls, Oken]{Chrisna Aing, Sarah Halls, Kiva Oken}
\institute{Carleton College}
\usepackage{graphpap,graphics}
\includeonlyframes{currentslide}

@@

\section{Exploring the data}
\begin{frame}{Effects of multiple observers}
	\begin{itemize}
		\item In 2003, three different observers
		\item Detection rates varied among the observers
		%include graph here
		\item Solution: use observer indicators as covariates for detection 
		probability
	\end{itemize}
\end{frame}

@@

%This slide for DNR only
\begin{frame}{Bayesian model specifics}
	\begin{itemize}
		\item $\psi,\theta \sim \text{Uniform}(0,1)$
		\item $p_{\text{obs}} \sim \text{Beta}(a,b)$
		\begin{itemize}
			\item $a, b \sim \text{Gamma}(1,0.1)$
			\item E similar
			\item $p$ and $E$ had restricted distributions
		\end{itemize}
		\item Burn-in period 10,000 iterations
		\item Covariates: observer, days since snow
	\end{itemize}
\end{frame}

@@

\begin{frame}{Modeling spatial correlation}
	\begin{itemize}
		\item Conditional autoregressive (CAR) models
		\item Allows extreme values to influence estimates for neighboring sites
		\item Placed CAR prior on $\theta$, track-laying probability
		\item With CAR model, each site has a unique estimate of $\theta$
	\end{itemize}
\end{frame}

%DNR only slide
\begin{frame}{CAR model specifics}
	\begin{itemize}
		\item $\tau^2 =$ conditional variance
		\item $(\tau^2)^{-1} \sim \text{Gamma}(0.5,0.0005)$
		\item $\text{logit}(\theta_s)=\alpha_0+\alpha_{1_s}$
		\item $\alpha_0 \sim$ Improper uniform, $\alpha_1 \sim$ Gaussian CAR
	\end{itemize}
\end{frame}

%DNR slide
\begin{frame}[label=currentslide]{Testing the CAR model}
	\begin{itemize}
		\item Simulated data with correlation in $\theta$
		\begin{itemize} 
			\item $P(\text{track in site }i|\text{no track in site }i-1)=\theta$
			\item $P(\text{track in site }i|\text{track in site }i-1)=\theta+
			\alpha$
		\end{itemize}
		\item CAR model did not converge
		\item Simulated data more independent than actual data (from $\hat{S}$)
		\item Added similar correlation in $\psi$
		\begin{itemize}
			\item Data converged more often
			\item Independence more similar to that of actual data
		\end{itemize}
	\end{itemize}
\end{frame}

%Carleton slide
\end{document}

>>>>

\documentclass{beamer}
\usetheme{}
\title{Using occupancy theory and Bayesian models to estimate otter populations in southeastern Minnesota}
\author[Aing, Halls, Oken]{Chrisna Aing, Sarah Halls, Kiva Oken}
\institute{Carleton College}
\usepackage{graphpap,graphics}
%\includeonlyframes{currentslide}

@@

\section{Exploring the data}
\begin{frame}{Effects of multiple observers}
	\begin{itemize}
		\item In 2003, three different observers
		\item Detection rates varied among the observers
	\end{itemize}
	\begin{center}
		\resizebox{5cm}{!}{\includegraphics{../observerPlots.pdf}}
	\end{center}
	\begin{itemize}
		\item Solution: use observer indicators as covariates for detection 
		probability
	\end{itemize}
\end{frame}

@@

%This slide for DNR only
\begin{frame}{Bayesian model specifics}
	\begin{itemize}
		\item $\psi,\theta \sim \text{Uniform}(0,1)$
		\item $p[{\text{obs}] \sim \text{Beta}(a,b)$
		\begin{itemize}
			\item $a, b \sim \text{Gamma}(1,0.1)$
			\item E similar
			\item $p$ and $E$ had restricted distributions
		\end{itemize}
		\item Burn-in period 10,000 iterations
		\item Covariates: observer, days since snow
	\end{itemize}
\end{frame}

@@

\begin{frame}{Modeling spatial correlation}
	\begin{itemize}
		\item Conditional autoregressive (CAR) models
		\item Allows extreme values to influence estimates of neighboring sites
		\item Placed CAR prior on $\theta$, track-laying probability
		\item With CAR model, each site has a unique estimate of $\theta$
	\end{itemize}
\end{frame}

%DNR only slide
\begin{frame}{CAR model specifics}
	\begin{itemize}
		\item $\tau^2 =$ conditional variance
		\item $(\tau^2)^{-1} \sim \text{Gamma}(0.5,0.0005)$
		\item $\text{logit}(\theta_s)=\alpha_0+\alpha_{1_s}$
		\item $\alpha_0 \sim$ Improper uniform
		\item $\alpha_1 \sim$ Gaussian CAR
	\end{itemize}
\end{frame}

%DNR slide
\begin{frame}{Testing the CAR model}
	\begin{itemize}
		\item Simulated data with correlation in $\theta$
		\begin{itemize} 
			\item $P(\text{track in site }i|\text{no track in site }i-1)=\theta$
			\item $P(\text{track in site }i|\text{track in site }i-1)=\theta+
			\alpha$
		\end{itemize}
		\item CAR model did not converge
		\item Simulated data more independent than actual data (from $\hat{S}$)
		\item Added similar correlation in $\psi$
		\begin{itemize}
			\item CAR model converged more often
			\item Independence more similar to that of actual data
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Testing CAR model, continued}
	\begin{itemize}
		\item With correlated data, no constant $\psi$.
		\item To find true $\psi$ from simulated data, took proportion of plots 
		occupied
		\item When CAR model converged, estimates from CAR and standard models 
		similar for the same data
	\end{itemize}
\end{frame}

%Carleton slide goes here
\section{Discussion}
\begin{frame}{Obtaining convergence on the data}
	\begin{itemize}
		\item All of 2003 data converged with standard model
		\item Nothing in 2004 converged
		\begin{itemize}
			\item Removed false detection, 402m in 2004 converged
		\end{itemize}
		\item CAR model never converged with $E$. Without $E$, poor 
		estimates for $\psi$
	\end{itemize}
\end{frame}

\begin{frame}{Summary and interpretation of estimates}
	\begin{itemize}
		\item Estimates for all parameters increase with plot size
		\item Possible hypotheses...
		\item Used data from all plots, not alternating plots
		\begin{itemize}
			\item More data more useful than more independence
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Recommended sampling practices}
	\begin{itemize}
		\item Investigated effects of changing the sampling variables
		\item In order of decreasing importance, maximize:
		\begin{enumerate}
			\item Number of consecutive days after snow event sampled
			\item Number of snow events
			\item Number of observers
		\end{enumerate}
		\item All conclusions come from regular data
	\end{itemize}
\end{frame}

\begin{frame}{Acknowledgments}
	\begin{itemize}
		\item Bob Dobrow, adviser
		\item John Fieberg, MN DNR
		\item John Erb, MN DNR
	\end{itemize}
\end{frame}

\begin{frame}{Questions?}
\end{frame}

\end{document}
